Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-05 paper.tex
add abstract, intro, and some results

\documentclass[]{article}
\usepackage[round]{natbib}

@@

% local definitions
\newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
\newcommand{\bmhcomment}[1]{{\textcolor{blue}{BMH: #1}}}
\newcommand{\aprcomment}[1]{{\textcolor{purple}{APR: #1}}}
\newcommand{\tdwcomment}[1]{{\textcolor{orange}{TDW: #1}}}

@@

\abstract{
A very simple template for an article class document.
}

@@

The intro..

@@

@@

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
\caption{\textbf{The first figure}.
    The geographic and genetic diversity of populations across Africa.
}
\label{fig:1}
\end{center}
\end{figure}

@@

\begin{figure}[ht]
\begin{center}
    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
\caption{\textbf{The third main figure}.
    A placeholder - validation (Relate and cSFS).
}
\label{fig:r3}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
\caption{\textbf{The fourth main figure}.
    A placeholder - predictions (FST and/or f4).
}
\label{fig:r4}
\end{center}
\end{figure}

@@

\bibliographystyle{plainnat}
\bibliography{paper}
\end{document}

>>>>

\documentclass[]{article}
\usepackage[super,numbers]{natbib}

@@

% local definitions
\newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
\newcommand{\bmhcomment}[1]{{\textcolor{blue}{BMH: #1}}}
\newcommand{\aprcomment}[1]{{\textcolor{magenta}{APR: #1}}}
\newcommand{\tdwcomment}[1]{{\textcolor{cyan}{TDW: #1}}}

@@

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135ka, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral Homo spp. populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
\sgcomment{IF we were not afraid of the racists, we could write something like:
"Despite the genetic similarity between these populations, 4 to 10\% of random
genetic differences among [modern human populations] can be attributed to this
early genetic drift.}
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar. We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
distributed across Africa, and are particularly well represented in the
northern, eastern and southern parts of the continent. Similarly, fossil crania
such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
anatomically derived \emph{Homo sapiens} features were also present across the
continent during this period. It has been difficult to reconcile these lines of
evidence with evidence from genomics, which have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population. It is unclear whether fossil specimens and archaeological sites
represent populations which contributed to our ancestors as population
precedents, or were local “dead-ends” from which contemporary \emph{Homo
sapiens} do not descend. Recently, synthetic attempts to reconcile genetic and
paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of Homo sapiens beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data. However, new
population genetic inference tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with modern humans in Eurasia, several recent articles
have shown that introducing an archaic ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific archaic remains, and the possibility of
finding ancient DNA evidence (e.g., \citet{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic admixture model, leaving out other plausible models
(Figure~\ref{fig:supp-possible-models} and \citet{Henn2018-rf}). Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African ancestry. While ancient DNA from
Eurasia has helped us understand early human history outside of Africa, there
is no comparably ancient DNA to elucidate early history in Africa.

Here, we therefore aim to discriminate among a broader set of demographic
models by interrogating the genomes of contemporary populations. We take as our
starting point 4 classes of models (single population expansion, single
    population expansion with regional persistence, archaic admixture, and
multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
genomes from southern, eastern, and western Africa as well as Eurasia. By
including geographically and genetically diverse populations across Africa, we
infer demographic models that explain more aspects of genetic diversity in more
populations than previously reported. These analyses confirm the inadequacy of
tree-like models and provide an opportunity to directly evaluate a wide range
of alternative models.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (MSL from Sierre Leone
\citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz (recent descendents of a
hunter-gatherer group from Ethiopia \citep{Gurdasani2015-qy,Gopalan2019-wd}),
and Eastern African agriculturalists (Amhara and Oromo from Ethiopia
\citep{Gurdasani2015-qy}). The Amhara and Oromo populations, despite speaking
distinct Afro-Asiatic languages, are highly genetically similar
\citep{Pagani2015-pz,Gopalan2019-wd} and thus the two groups were combined for
a larger sample size (Figure~\ref{fig:1}). We also included the British
(GBR) from the 1000 Genomes Project in our demographic models as a
representative source of back-to-Africa gene flow and recent colonial admixture
in South Africa. Finally, we added a high-coverage ancient Neanderthal genome
from Vindija Cave, Croatia \citep{Prufer2017-kk} to account for archaic gene
flow from Neanderthals in non-Africans and gauge the relative time depth of
divergence, assuming Neanderthals diverged 550ka from a common stem. For each
population, we computed low-order allele frequency and linkage disequilibrium
(LD) statistics that are well suited for both low- and high-coverage genomes
\citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a maximum-likelihood inference
framework, we then fit to these statistics a family of parameterized
demographic models that involve population splits, size changes, continuous and
variable migration rates, and punctuated admixture events, to learn about the
nature of population structure over the past million years.

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
\caption{\textbf{Geography and genetic diversity within Africa}.
    \aprcomment{In the models presented below, we focus on the Nama, 
    Mende (MSL), Gumuz, and Oromo and Amhara.}
}
\label{fig:1}
\end{center}
\end{figure}

@@

We began with a model of geographic expansion from a single ancestral source
followed by migration between populations
(Figure~\ref{fig:supp-possible-models}A), without allowing for contribution
from an archaic African lineage. As expected \citep{Ragsdale2019-nt}, this
first model was a poor fit to the data qualitatively
(Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
($LL$) $\approx -189,000$, Table~\ref{tab:supp-single-origin}). We next explored a
suite of models in which population structure predates the differentiation of
contemporary groups, including models allowing for ancestral reticulation
(Figure~\ref{fig:supp-possible-models}B), archaic admixture
(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
(Figure~\ref{fig:supp-possible-models}D).

Regardless of the model choice for early epochs, inference of recent human
demographic history from two thousand \aprcomment{why two thousand, instead of
just ``up to 150ka''?} to 150ka was remarkably robust. The earliest divergence
among contemporary sampled human populations differentiates the southern
African Nama from other African groups between 110--135ka, with low to moderate
levels of subsequent gene flow \aprcomment{(Table 1)}. In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$150ka. We conclude that, regardless of model
specification, geographic patterns of \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems \aprcomment{(Figure 3)}.

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (Supp.
Methods). The time of divergence of Western and Eastern African populations was
set to 60ka, just prior to the split of Eurasians and East Africans 50ka. Both
models predict relatively high subsequent gene flow between the two regions
($m=2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by ADMIXTURE
\aprcomment{cite} (Figure~\ref{fig:1}).

@@

@@

\begin{figure}[ht]
\begin{center}
    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
\caption{\textbf{The third main figure}.
    A placeholder - validation (Relate and cSFS).
}
\label{fig:3}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
\caption{\textbf{The fourth main figure}.
    A placeholder - predictions (FST and/or f4).
}
\label{fig:4}
\end{center}
\end{figure}

@@

\bibliographystyle{unsrtnat}
\bibliography{paper}
\end{document}

### 2022-01-05 supplement.tex
add abstract, intro, and some results

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{bm}

@@

\newcommand{\E}{\mathbb{E}}

@@

\sgcomment{To be expanded. TODO: Brenna}
\begin{itemize}
    \item From the merged dataset of the African Diversity Reference Panel and
        1000 Genomes phase 3 data, subsampled to population included in this study
    \item West Africa: Mende from Sierra Leone (MSL); South Africa: newly sequenced 
        Nama; East Africa: Gumuz (traditionally hunter-gatherer with low levels of 
        Eurasian admixture), Oromo and Amhara (combined as traditionally 
        agriculturalists with large proportion of back-to-Africa Eurasian ancestry); 
        Eurasian: British (GBR)
    \item Combined with the high coverage Vindija neanderthal genome
    \item For running Relate, we used a larger set of populations. We kept all 
        African 1000 Genomes populations, along with the Nama, Gumuz, Oromo, and
        Amhara, as well as 4 Eurasian populations from 1000 Genomes: GBR, CEU, PJL,
        and CHB
\end{itemize}

@@

All analyses presented in this work focus on biallelic single nucleotide
polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the moments-LD
analysis, we focused on intergenic locations because these appear less affected
by natural selection compared to both synonymous and nonsynonymous variation
\citep{Ragsdale2018-dd}. To enable comparison with Neanderthal DNA, we excluded
regions for which the Vindija Neanderthal sample had less than 100 contiguous
base pairs.

@@

We compared single- and two-locus statistics in the data to predictions based
on detailed demographic models. Model predictions were obtained using
recursions described in \citet{Ragsdale2019-nt} and implemented in the software
\emph{moments.LD} (\url{https://bitbucket.org/simongravel/moments/src/main/}).
The model computes expected patterns of single-nucleotide pairwise diversity
and linkage disequilibrium as a function of recombination distance between
variants within and across populations, under the assumption of neutrality.

@@

The approach from \citet{Ragsdale2020-nz} provides unbiased estimates of the LD
statistics considered here, which can be directly compared to expectations for
those same statistics computed from the multi-population Hill-Robertson system.
Estimated statistics for populations with smaller sample sizes have greater
uncertainty, which is largely accounted for by computing variances and
covariances via bootstrap, assuming our samples are unrelated and the
population is randomly mating.

\sgcomment{I don't think that this is true if our bootstrap is
over genomic regions. In an infinite genome with a tiny sample size, we would
estimate no uncertainty...  Or at least, it is only true if the Neandertal form
a truly randomly mating population...  }

@@

moments-LD uses a composite likelihood approach to simultaneously fit relative
pairwise diversity and LD statistics over a range of recombination distances.
Likelihoods were computed independently for pairwise diversity and each
recombination distance bin using a multivariate Guassian likelihood function as
described in Section \ref{sec:computing-stats}. These were multiplied across
bins and the single set of heterozygosity statistics following the approach
detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
rounds of optimization, alternating between the \emph{optimize\_log\_fim},
\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
parameter space and hone in on the best fit parameters. Initial guesses for
parameters were chosen from demographically plausible starting points and then
perturbed to explore space, using gradient descent (on the log of the
parameters). The best fits from these initial rounds of optimization were then
chosen as the starting points for optimization using the Powell and/or the
L-BFGS-B methods (as implemented in the SciPy optimization package
\cite{Virtanen2020-kr}). This process was repeated with alternating
optimization methods until the best fit parameter set converged consistently.

@@

We used Relate version 1.0.16 \citep{Speidel2019-nj} to reconstruct genome-wide
gene genealogies using a combined set of 1000 Genomes and African Diversity
Reference Panel datasets, retaining all AFR-labeled populations and GBR, CEU,
PJL, and CHB from the 1000 Genomes panel and the Nama, Gumuz, Oromo, and Amhara
from ADRP. We used all autosomes and applied the 1000 Genomes Phase 3 strict
mask, we used an ancestral sequence determined by a 6-primate alignment
(human\_ancestor\_GRCh37\_e59), and we used the HapMap II combined
recombination map, all in GRCh37 coordinates. We assumed a generation time of
29 years, a mutation rate of 1.25e-8, and followed the standard pipeline
described in the Relate documentation.

From the reconstructed gene genealogies, we computed coalescence rates within
and between populations using Relate’s function RelateCoalescentRate. This
allows for an estimate of the instantaneous inverse coalescence rate (IIRC) for
samples drawn within each population (the inverse of which is often interpreted
as the effective population size history), and the relative cross coalescence
rates between pairs of populations (which are commonly used to estimate
divergence times).

@@

@@

@@

@@

\subsection{Conditional SFS}

\subsection{Relate curves from inferred models}

\subsection{Distribution of deep branch affinities to Neanderthal sequence}

@@

\subsection{Re-inference of an IM model from simulated data}

@@

\addcontentsline{toc}{section}{Supporting tables}
\section*{Supporting tables}

@@

\addcontentsline{toc}{section}{Supporting figures}
\section*{Supporting figures}

>>>>

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{xspace}

@@

\newcommand{\E}{\mathbb{E}}
\newcommand{\moments}{\texttt{moments}\xspace}
\newcommand{\Relate}{\texttt{Relate}\xspace}
\newcommand{\msprime}{\texttt{msprime}\xspace}
\newcommand{\demes}{\texttt{demes}\xspace}

@@

\sgcomment{To be expanded. TODO: Brenna}
\begin{itemize}
    \item From the merged dataset of the African Diversity Reference Panel and
        1000 Genomes phase 3 data, subsampled to population included in this study
    \item West Africa: Mende from Sierra Leone (MSL); South Africa: newly sequenced 
        Nama; East Africa: Gumuz (traditionally hunter-gatherer with low levels of 
        Eurasian admixture), Oromo and Amhara (combined as traditionally 
        agriculturalists with large proportion of back-to-Africa Eurasian ancestry); 
        Eurasian: British (GBR)
    \item Combined with the high coverage Vindija neanderthal genome
    \item For running \Relate, we used a larger set of populations. We kept all 
        African 1000 Genomes populations, along with the Nama, Gumuz, Oromo, and
        Amhara, as well as 4 Eurasian populations from 1000 Genomes: GBR, CEU, PJL,
        and CHB
\end{itemize}

@@

All analyses presented in this work focus on biallelic single nucleotide
polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the
\moments-LD analysis, we focused on intergenic regions because these
appear less affected by natural selection compared to both synonymous and
nonsynonymous variation \citep{Ragsdale2018-dd}. To enable comparison with
Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
had less than 100 contiguous base pairs.

@@

We compared single- and two-locus statistics in the data to predictions based
on detailed demographic models. Model predictions were obtained using
recursions described in \citet{Ragsdale2019-nt} and implemented in the software
\moments (\url{https://bitbucket.org/simongravel/moments/src/main/}).
The model computes expected patterns of single-nucleotide pairwise diversity
and linkage disequilibrium as a function of recombination distance between
variants within and across populations, under the assumption of neutrality.

@@

The approach from \citet{Ragsdale2020-nz} provides unbiased estimates of the LD
statistics considered here, which can be directly compared to expectations for
those same statistics computed from the multi-population Hill-Robertson system.
Estimated statistics for populations with smaller sample sizes have greater
uncertainty, which is accounted for by computing variances and covariances via
bootstrap over genomic regions, assuming that there is no correlation between
distant genomic regions (i.e., no long-range LD).

@@

\moments-LD uses a composite likelihood approach to simultaneously fit
relative pairwise diversity and LD statistics over a range of recombination
distances.  Likelihoods were computed independently for pairwise diversity and
each recombination distance bin using a multivariate Guassian likelihood
function as described in Section \ref{sec:computing-stats}. These were
multiplied across bins and the single set of heterozygosity statistics
following the approach detailed in \citet{Ragsdale2019-nt}. For each model
tested, we ran multiple rounds of optimization, alternating between the
\emph{optimize\_log\_fim}, \emph{optimize\_log\_powell}, and
\emph{optimize\_lbfgsb} methods to explore parameter space and hone in on the
best fit parameters. Initial guesses for parameters were chosen from
demographically plausible starting points and then perturbed to explore space,
using gradient descent (on the log of the parameters). The best fits from these
initial rounds of optimization were then chosen as the starting points for
optimization using the Powell and/or the L-BFGS-B methods (as implemented in
the SciPy optimization package \cite{Virtanen2020-kr}). This process was
repeated with alternating optimization methods until the best fit parameter set
converged consistently.

@@

We used \Relate version 1.0.16 \citep{Speidel2019-nj} to reconstruct genome-wide
gene genealogies using a combined set of 1000 Genomes and African Diversity
Reference Panel datasets, retaining all AFR-labeled populations and GBR, CEU,
PJL, and CHB from the 1000 Genomes panel and the Nama, Gumuz, Oromo, and Amhara
from ADRP. We used all autosomes and applied the 1000 Genomes Phase 3 strict
mask, we used an ancestral sequence determined by a 6-primate alignment
(human\_ancestor\_GRCh37\_e59), and we used the HapMap II combined
recombination map, all in GRCh37 coordinates. We assumed a generation time of
29 years, a mutation rate of 1.25e-8, and followed the standard pipeline
described in the \Relate documentation.

From the reconstructed gene genealogies, we computed coalescence rates within
and between populations using \Relate’s function \texttt{RelateCoalescentRate}.
This allows for an estimate of the instantaneous inverse coalescence rate
(IIRC) for samples drawn within each population (the inverse of which is often
interpreted as the effective population size history), and the relative cross
coalescence rates between pairs of populations (which are commonly used to
estimate divergence times).

@@

Here we focus on four parameterizations of early human history: (1) a “single
origin” with recent expansion, (2) primary and second stem populations with
“continuous migration”, (3) two stem populations that do not exchange migrants
before merging to form regional African populations, referred to as a “merger
without stem migration”, and (4) two stem populations that are allowed to
exchange migrants before merging to form regional African populations, referred
to as a “merger with stem migration”. Given the best fit parameters for each
model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
simulate sampled individual genomes to match the data used in inference (see
simulation details below). Using these simulations, we compared additional
statistics and inferences between simulated and real data that were not used in
the original optimization. These comparisons serve as independent validations
of our inferred demographic models by assessing how well those models fit
aspects of the data that have been used in other studies as evidence of
population structure or archaic admixture from an unidentified hominin species
within Africa.

@@

The four inferred demographic models (single origin, continuous migration,
merger without stem migration, and merger with stem migration) were converted
to \demes format (Gower, Ragsdale, \emph{et al.}, in prep) and used as
the input demography for \msprime simulations. We sampled the same
number of individuals from each population as used in the inference (Nama: 44,
Mende: 85, Gumuz: 23, combined Oromo/Amhara: 46, British: 91, and Vindija
Neanderthal: 1). We assumed a generation time of 29 years, a per-base mutation
rate of $1.25\times10^{-8}$, a per-base recombination rate of $1\times10^{-8}$
(corresponding to 1cM/Mb), and we simulated 100 replicates with sequence
lengths of 50Mb per replicate. We computed the conditional site frequency
spectrum (cSFS) conditioned on observing the derived variant in the Neanderthal
sample, and reconstructed gene genealogies using \Relate to obtain coalescence
rate trajectories and distributions of deep branches passing through the
Neanderthal branch. Each analysis followed the same procedure as described
above in the analysis of the real data.

@@

We compared the observed cSFS to model predictions for the Nama, Mende, and
Gumuz populations. The cSFS from data is sensitive to ancestral allele
misidentification, but our simulations assume perfect knowledge of the
ancestral allele. To investigate the role of misidentification, we compared
predicted and observed cSFS for transitions and transversions separately as
well as jointly.

For each comparison, we fit a misidentification parameter $p_{misid}$ as well
as a scaled mutation rate $\theta$. The misidentification parameter is the
fraction $p_{misid}$ of loci that have the ancestral state misidentified.
Accounting for $p_{misid}$ mixes the predicted joint population SFS with a
small amount of the ancestral-allele-reversed SFS (i.e. for a two-population
joint SFS, $SFS_{misid}(i, j)=(1-p_{misid})*SFS(i, j) + p_{misid}*SFS(n-i,
m-j)$), where $n$ and $m$ are the sample sizes from the two populations. The
cSFS for the first population (conditioning on the second) is found by
projecting the sample size to $(n, 1)$ and taking $cSFS=SFS(:,1)$. $\theta$ and
$p_{misid}$ were chosen to maximize a likelihood assuming that each observed
entry in the cSFS is drawn from a Poisson distribution \citep{Sawyer1992-rt}.

@@

\subsection{Alternative statistics for four models}

\subsubsection{Conditional SFS}

The single-origin model fit the cSFS relatively poorly across the three
populations compared to the models that allow structure between stem
populations (Fig.~\ref{fig:supp-csfs-single-origin}). The model that best fit
the cSFS was the merger-with-stem-migration model, which also provided the best
fit to pairwise diversity and LD statistics in the initial optimization. For
the Mende and Gumuz, the merger-with-stem-migration model fits the data very
well, while it underestimates rare variants in the Nama cSFS
(Fig.~\ref{fig:supp-csfs-merger-with-stem-migration}). Models learned from the
two-locus statistics provided a better qualitative description of the cSFS than
models previously fit directly to the cSFS  (see, e.g., figure 1 in
\citet{Durvasula2020-td}). We found that the inferred ancestral
misidentification rates were concordant across populations. The inferred rates
of ancestral misidentification were about twice as high for transitions
relative to transversions as expected, qualitatively, from the higher
transition mutation rate \citep{Hernandez2007-mf}.

\subsubsection{Relate curves from inferred models}

\aprcomment{TODO}

\subsubsection{Distribution of deep branch affinities to Neanderthal sequence}

\aprcomment{TODO}

@@

Prior studies primarily relied on the mutation clock in order to date events
such as population divergence, and uncertainty about the mutation rate
translates in uncertainty in timing estimates (e.g., \citet{Moorjani2016-qj}).
Recent whole genome sequencing studies from non-pathological pedigrees have
typically estimated mutation rates to be $1.2-1.3\times10^{-8}$ per site per
generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
approximates the rate predicted by a mutational model from Gnomad
\citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
demographic inference in this study. However, \moments differs from other
approaches by relying on a recombination clock, which has been argued to be
more robust to estimation than a mutational clock \citep{Moorjani2016-ur}. To
fit patterns of diversity which were inferred using the recombination map here,
mutation rates on the order of $2\times10^{-8}$ would be needed.
\aprcomment{This discrepancy between the mutation clock and recombination
clock\dots}

\subsection{Re-inference of IM models from simulated data}

\aprcomment{TODO}

@@

\addcontentsline{toc}{section}{Supporting tables and figures}
\section*{Supporting tables and figures}

@@
